Providing A Unified User Environment for Canada National Advanced Computing Centers

### 2019-02-12 extended-abstract.tex
added labels and updated references to sections

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}

%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Providing A Unified User Environment for Canada’s National Advanced Computing Centers}

@@

The first principle is that software packages should be available on every cluster (unless not permitted because of
license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are
a few solutions that could be used, and we will discuss them in Section 1. The solution we opted for is Cern Virtual
Machine File System, or CVMFS\cite{CVMFS}.

The second principle is that the software environment should be decoupled from the underlying operating system. When we
first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, there was
interest to support some clusters running CentOS 6. Moreover, not every CentOS 7 system has the same set of system
packages installed. Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual
machines running any Linux distribution. In order to use the unified software environment on virtual machines and
multiple clusters, we needed a compatibility layer. The solution we initially deployed is Nix \cite{Nix}, although we
are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will
discuss these options in Section 2.

@@

The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable
with a very large number of software packages installed. All of our clusters were using a system of environment modules
\cite{Modules1991,Modules1996}. However, the traditional systems of modules become unwieldy to handle at the scale of
thousands of installed modules. Using a hierarchical module tree, which is possible in a newer implementation of
environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets
the third and fourth principles in Section 3.

\section{Distribution mechanism}
Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or
GPFS. Anyone who has been working in the field of high performance computing for a number of years knows that parallel
file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications. In
addition to that problem, this solution is also a single point of failure, since any failure of that shared file system
will block the vast majority of jobs. Another solution for a single cluster is to install binary packages on each node,
using a mechanism such as RPM. This solution can however lead to diverging configurations on different nodes, and
require an additional layer to keep the configurations synchronized.

@@

\subsection{CVMFS structure}
CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called
the stratum 0. This is the injection point for new software packages that are to be deployed on the cluster. This
server is never accessed directly by the end users or the clusters. This means that if it has a failure, the only
consequence is that one can not inject new files until the server is back.

@@

\section{Compatibility layer}
In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost
everything, down to the C library. The only exclusions that we do not include are the Linux kernel, kernel modules,
drivers, and anything that is to be run as a privileged user.

\subsection{Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility.
Therefore, the recipes attempt very hard and mostly succeed to not depend on the underlying operating system at all.
Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its
dependencies. This ensures very strong reproducibility, but it also means that updating a package in place is virtually
impossible. While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or
security issues, which is made very hard by the design choices in Nix. At the time of writing this paper, we are still
using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.

\subsection{Gentoo Prefix}

\section{Scientific applications}

\subsection{Compiled software packages}
The majority of our software stack is made up of scientific software that are performance critical. This implies that
they cannot be installed in the compatibility layer described above, because different clusters support different sets
of CPU instructions. To install these packages, we use a package manager called EasyBuild
\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.

@@

\subsection{Handling multiple CPU architectures}
In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different
directory and module tree based on the architecture requested. There is also a wide selection of software packages that
are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that
are not performance critical, or packages that consists of source files only.

@@

\subsection{Packages for scripting languages}
Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl
scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as
modules. In the case of Python, we are building Python wheels \cite{Wheels} against our software stack. Wheels are
binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by
our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good
performance.

\subsection{Licensed software packages}
Compute Canada also has agreements with a number of software vendors or developers of software packages that require a
license. In these cases, we install the package using the same tools, but it is installed on a separate CVMFS
repository which is only accessible to Compute Canada clusters.

\section{Other benefits}
Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any
virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas
of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this
portable software environment to expose a cluster environment directly through a browser, via JupyterHub.

\section{Acknowledgments}

We thank the following organizations and individuals for sharing their time and expertise:
\begin{itemize}
	\item All of the Compute Canada team members who make it possible to manage such a large software stack
	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and
debugging for Lmod
	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions
regarding EasyBuild
	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this
project
	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS
\end{itemize}

>>>>

%
% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}
%\usepackage[french]{babel}
%\selectlanguage{french}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
%
% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Providing A Unified User Environment for Canada National Advanced Computing Centers}

@@

The first principle is that software packages should be available on every cluster (unless not permitted because of
license or hardware reasons). This means that we need a scalable resilient multi-site distribution mechanism. There are
a few solutions that could be used, and we will discuss them in Section~\ref{sec:Distribution_mechanism}. The solution
we opted for is Cern Virtual Machine File System, or CVMFS\cite{CVMFS}.

The second principle is that the software environment should be decoupled from the underlying operating system. When we
first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, there was
interest to support some clusters running CentOS 6. Moreover, not every CentOS 7 system has the same set of system
packages installed. Finally, Compute Canada also operates several OpenStack clouds, on which users can launch virtual
machines running any Linux distribution. In order to use the unified software environment on virtual machines and
multiple clusters, we needed a compatibility layer. The solution we initially deployed is Nix \cite{Nix}, although we
are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement because of issues we encountered with Nix. We will
discuss these options in Section~\ref{sec:Compatibility_layer}.

@@

The fourth principle is that the interface presented to users should be as familiar as possible, and remain manageable
with a very large number of software packages installed. All of our clusters were using a system of environment modules
\cite{Modules1991,Modules1996}. However, the traditional systems of modules become unwieldy to handle at the scale of
thousands of installed modules. Using a hierarchical module tree, which is possible in a newer implementation of
environment modules called Lmod \cite{Lmod}, helps address this problem. We will discuss how our implementation meets
the third and fourth principles in Section~\ref{sec:Scientific_applications}.

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
Installing software packages on a single cluster can be done through a shared file system, such as NFS, Lustre, or
GPFS. Anyone who has been working in the field of high performance computing for a number of years knows that parallel
file systems are often a bottleneck, especially for the IO pattern caused by the action of launching applications. In
addition to that problem, this solution is also a single point of failure, since any failure of that shared file system
will block the vast majority of jobs. Another solution for a single cluster is to install binary packages on each node,
using a mechanism such as RPM. This solution can however lead to diverging configurations on different nodes, and
require an additional layer to keep the configurations synchronized.

@@

\subsection{CVMFS structure}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, which is called
the stratum 0. This is the injection point for new software packages that are to be deployed on the cluster. This
server is never accessed directly by the end users or the clusters. This means that if it has a failure, the only
consequence is that one can not inject new files until the server is back.

@@

\section{Compatibility layer}
\label{sec:Compatibility_layer}
In order to detach the software stack from the underlying operating system, we need a layer that allows to build almost
everything, down to the C library. The only exclusions that we do not include are the Linux kernel, kernel modules,
drivers, and anything that is to be run as a privileged user.

\subsection{Nix}
\label{sub:Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). Its core principle is reproducibility.
Therefore, the recipes attempt very hard and mostly succeed to not depend on the underlying operating system at all.
Every software built with Nix is installed inside a folder named by the hash of its recipe and that of its
dependencies. This ensures very strong reproducibility, but it also means that updating a package in place is virtually
impossible. While reproducibility is a major consideration in advanced computing, we sometimes need to fix bugs or
security issues, which is made very hard by the design choices in Nix. At the time of writing this paper, we are still
using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix described next.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}

\section{Scientific applications}
\label{sec:Scientific_applications}

\subsection{Compiled software packages}
\label{sub:Compiled_software_packages}
The majority of our software stack is made up of scientific software that are performance critical. This implies that
they cannot be installed in the compatibility layer described above, because different clusters support different sets
of CPU instructions. To install these packages, we use a package manager called EasyBuild
\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016}.

@@

\subsection{Handling multiple CPU architectures}
\label{sub:Handling_multiple_CPU_architectures}
In order to handle multiple CPU architectures, we have configured EasyBuild to install software packages in a different
directory and module tree based on the architecture requested. There is also a wide selection of software packages that
are installed in an architecture-independent fashion. This is the case, for example, for binary packages, packages that
are not performance critical, or packages that consists of source files only.

@@

\subsection{Packages for scripting languages}
\label{sub:Packages_for_scripting_languages}
Whenever a software package supports the installation of subpackages, as is the case with the Python, R, or Perl
scripting languages, we leverage the existing package manager of that language, rather than installing subpackages as
modules. In the case of Python, we are building Python wheels \cite{Wheels} against our software stack. Wheels are
binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by
our team, and linked against our optimize software stack, we ensure that they are both easy to installed and offer good
performance.

\subsection{Licensed software packages}
\label{sub:Licensed_software_packages}
Compute Canada also has agreements with a number of software vendors or developers of software packages that require a
license. In these cases, we install the package using the same tools, but it is installed on a separate CVMFS
repository which is only accessible to Compute Canada clusters.

\section{Other benefits}
\label{sec:Other_benefits}
Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any
virtual machine on a cloud. For example, for training events, we have been using it to provision small sized replicas
of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this
portable software environment to expose a cluster environment directly through a browser, via JupyterHub.

\section{Acknowledgments}
\label{sec:Acknowledgments}
We thank the following organizations and individuals for sharing their time and expertise:
\begin{itemize}
	\item All of the Compute Canada team members who make it possible to manage such a large software stack
	\item Robert McLay, from Texas Advanced Computing Center, for always providing quick and efficient support and
debugging for Lmod
	\item Kenneth Hoste from Ghent University, and Åke Sandgren from Umeå University, for many discussions
regarding EasyBuild
	\item Kuang Chung Chen and Bryan Caron, from McGill University, for initiating the groundwork that lead to this
project
	\item Dave Dykstra, Jakob Blomer, and the CernVM project team for support and development of CVMFS
\end{itemize}
